---
title: "Fitting The Cauchy Distribution"
author: "Michael Betancourt"
date: "January 2018"
output:
  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA)
```

The Cauchy distribution has a notorious history of serving as a
counterexample to mathematical proofs in probability theory.  At
the same time it has been a consistent source of frustration for
statistical computation as well.  In this case study I review
various ways of implementing the Cauchy distribution, from the
nominal implementation to alternative implementations aimed at
ameliorating these difficulties, and demonstrate their relative
performance.

# The Cauchy Distribution

The Cauchy distribution seems innocent enough, with a density given by
the rational function,
$$
\pi(x) = \frac{1}{\pi \, s}
\frac{ s^{2} }{ (x - m)^{2} + s^{2}}.
$$
Here the location, $m$, describes a certain sense of location of the
distribution while the scale, $s$, describes a certain sense of its
width.

```{r}

c_light <- c("#DCBCBC")
c_light_highlight <- c("#C79999")
c_mid <- c("#B97C7C")
c_mid_highlight <- c("#A25050")
c_dark <- c("#8F2727")
c_dark_highlight <- c("#7C0000")

x <- seq(-10, 10, 0.001)
plot(x, dcauchy(x, location = 0, scale = 1), type="l", col=c_dark_highlight, lwd=2,
     main="", xlab="x", ylab="Probability Density", yaxt='n')
```

This density, however, encodes extremely long tails that place significant
probability mass at points far away from $x = m$.  We can see just how heavy
these tails are by comparing the quantile function of the standardized Cauchy
distribution to that of a standardized Gaussian distribution,

```{r}
x <- seq(0, 1, 0.001)
plot(x, qcauchy(x, location = 0, scale = 1), type="l", col=c_dark_highlight, lwd=2,
     main="", xlab="Probability", ylab="Quantile")
lines(x, qnorm(x, 0, 1), type="l", col=c_light_highlight, lwd=2)

text(x=0.9, y=250, labels="Cauchy", col=c_dark_highlight)
text(x=0.9, y=-50, labels="Normal", col=c_light_highlight)
```

$95\%$ of the mass of the Gaussian distribution is contained within a distance
of $x = 1.6$ from $x = m = 0$, where as we would have to go to  $x = 6.3$
to contain the same probability mass for the Cauchy distribution.  In order to
contain $99\%$ of the mass we would need to go to $x = 2.3$ for the Gaussian
distribution but all the way to $x = 31.8$ for the Cauchy distribution!

These extremely heavy tails yield all kinds of surprising behavior for the
Cauchy distribution.  All of the even moments of the distribution are infinite
whereas all of the odd moments are not well defined.  In order to avoid
pathological behavior we have to restrict ourselves to characterizing the
Cauchy distribution with quantiles.

The Cauchy distribution was once recommended as the default for weakly
informative priors, but the behavior of these heavy tails, especially the
weak containment of probability mass around the location, $ x = m$, proved
to be too ungainly.  Consequently we have since moved towards recommending
Gaussian distributions for weakly informative priors.  Still, because the
Cauchy arises as a component in more sophisticated prior distributions such
as the horseshoe and its generalizations, understanding how to best fit the
distribution remains important.

# The Nominal Implementation of the Cauchy Distribution

The heavy tails of the Cauchy distribution makes it notoriously difficult
to estimate its expectation values.  In particular Random Walk Metropolis, the
Metropolis-Adjusted Langevin Algorithm, and even static Hamiltonian Monte Carlo
all fail to provide accurately estimates for these expectations.  The problem
is that for accurate estimation any algorithm will have to explore the massive
extent of the heavy tails, but once out in those tails there most algorithms
have difficulty returning back to the bulk of the distribution around $m = 0$.

How does the dynamic Hamiltonian Monte Carlo method used in Stan fare?
It's easy enough to check, here with a product of fifty Cauchy distributions,

```{r}
writeLines(readLines("cauchy_nom.stan"))
```

Pushing the program through Stan,

```{r}
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

util <- new.env()
source('stan_utility.R', local=util)
source('plot_utility.R', local=util)

fit_nom <- stan(file='cauchy_nom.stan', seed=4938483,
                warmup=1000, iter=11000, control=list(max_treedepth=20))

util$check_all_diagnostics(fit_nom, max_depth=20)
```

After increasing the maximum treedepth to a pretty high value we see no
indications of pathological fitting behavior.  We do, however, see $\hat{k}$
warnings which indicate that the $x$ parameters probably don't have finite
means and variances and hence well-defined effective sample sizes!

Consequently we shouldn't try to estimate the means or the estimated effective
samples sizes of these parameters and instead focus on quantiles of each $x$
or the mean of the included indicator function, $I$.

Indeed Stan is able to recover the $5\%$, $50\%$, and $95\%$ quantiles of
each parameter quite accurately,

```{r}
util$plot_estimated_quantiles(fit_nom, "Nominal Parameterization")
```

The variable integration times in dynamic Hamiltonian Monte Carlo allow
extremely long trajectories once we're out in the tails of the Cauchy
distribution.  These trajectories very slowly but surely carry us back
into the bulk of the distribution no matter how far into the tails we
have have sojourned.  The dynamic nature of the trajectories is important
here -- once we are deep enough into the tails any static but finite
integration time will not be long enough to ensure that we return.

Considering that most algorithms fail to accurately fit the Cauchy
distribution, this is a pretty remarkable achievement for Stan.
That said, the long trajectories required for this accuracy come at a
pretty significant computational cost.  Consequently let's consider
alternative implementations of the Cauchy distribution that can be
fit with less computational resources.

# First Alternative Implementation

An interesting property of the Cauchy distribution is that it arises
as a scale mixture of Gaussian distributions,
$$
\text{Cauchy}(x \mid m, s)
= \int \mathrm{d} \tau \,
\text{Normal} \, \left(x \mid m, \tau^{-\frac{1}{2}} \right)
\, \text{Gamma} \, \left(\tau \mid \frac{1}{2}, \frac{s^{2}}{2} \right).
$$
In other words, if
$$
x_{a} \sim \text{Normal} \, \left(0, 1 \right)
$$
and
$$
x_{b} \sim \text{Gamma} \, \left(\frac{1}{2}, \frac{s^{2}}{2} \right)
$$
then
$$
x = m + \frac{ x_{a} }{ \sqrt{x_{b}} }
$$
follows a $\text{Cauchy}(m, s)$ distribution.  Note that I am using
the Stan conventions for the Normal and Gamma density functions.

This property suggests a parameter expansion approach to implementing
the Cauchy distribution where we fit $x_{a}$ and $x_{b}$ and then
derive $x$ deterministically.  Although this requires twice the
number of parameters, the resulting joint density is much more
concentrated than the Cauchy distribution and hence significantly
easier to fit.

```{r}
x <- seq(-3, 3, 0.05)
y <- seq(-9, 1, 0.05)

n_x <- length(x)
n_y <- length(y)
z <- matrix(nrow=n_x, ncol=n_y)
for (i in 1:n_x) for (j in 1:n_y)
  z[i, j] <- dnorm(x[i], 0, 1) * dgamma(exp(y[j]), 0.5, 1 / 0.5) * exp(y[j])

contour(x, y, z, levels=seq(0.05, 1, 0.05) * max(z), drawlabels=FALSE,
        main="First Alternative", xlab="x_a", ylab="log(x_b)",
        col=c_dark_highlight, lwd=2)
```

This alternative implementation is straightforward to write as a Stan program,

```{r}
writeLines(readLines("cauchy_alt_1.stan"))
```

and readily fit,

```{r}
fit_1 <- stan(file='cauchy_alt_1.stan', seed=4938483,
              warmup=1000, iter=11000)

util$check_all_diagnostics(fit_1)
```

There are no indications of pathological behavior with Stan's default
settings, although we are again warned to be careful about interpreting
expectations of the reconstructed Cauchy parameters.  That said, we once again
accurately recover the quantiles of each of those recovered Cauchy parameters,

```{r}
util$plot_estimated_quantiles(fit_1, "First Alternative")
```

# Second Alternative Implementation

This first alternative implementation immediately suggests a second.
We can avoid the division in the recovery of $x$ by giving $x_{b}$
and _inverse_ gamma distribution.  Here we let
$$
x_{a} \sim \text{Normal} \, \left(0, 1 \right)
$$
and
$$
x_{b} \sim \text{Inv-Gamma} \, \left(\frac{1}{2}, \frac{s^{2}}{2} \right)
$$
from which
$$
x = m + x_{a} \cdot \sqrt{x_{b}}
$$
will follow a $\text{Cauchy}(m, s)$ distribution.

Although this may seem like a small change, the division operator and its
derivatives are significantly slower to evaluate than the multiplication
operator and its derivatives which can yield nontrivial performance
improvements.

This small change yields a joint density that mirrors the joint density
of the first alternative implementation and its pleasant geometry.

```{r}
x <- seq(-3, 3, 0.05)
y <- seq(-1, 9, 0.05)

n_x <- length(x)
n_y <- length(y)
z <- matrix(nrow=n_x, ncol=n_y)
for (i in 1:n_x) for (j in 1:n_y)
  z[i, j] <- dnorm(x[i], 0, 1) * dgamma(exp(-y[j]), 0.5, 1 / 0.5) * exp(-y[j])

contour(x, y, z, levels=seq(0.05, 1, 0.05) * max(z), drawlabels=FALSE,
        main="Second Alternative", xlab="x_a", ylab="log(x_b)",
        col=c_dark_highlight, lwd=2)
```

The corresponding Stan program is given by a small tweak

```{r}
writeLines(readLines("cauchy_alt_2.stan"))
```

and the fit

```{r}
fit_2 <- stan(file='cauchy_alt_2.stan', seed=4938483,
              warmup=1000, iter=11000)

util$check_all_diagnostics(fit_2)
```

proceeds with no issues save for the warnings about the recovered Cauchy
parameters.  As before, we accurately recover the quantiles of each Cauchy
distributed component in our model,

```{r}
util$plot_estimated_quantiles(fit_2, "Second Alternative")
```

# Third Alternative Implementation

The final alternative implementation that we will consider utilizes the
inverse cumulative distribution function of the Cauchy distribution.
In particular, if
$$
\tilde{x} \sim \text{Uniform} \, \left(0, 1 \right)
$$
then
$$
x = m + s \cdot \tan \, \left(\pi \left(\tilde{x} - \frac{1}{2} \right) \right)
$$
follows a $\text{Cauchy}(m, s)$ distribution.

The latent density of $\text{logit}(\tilde{x})$, which is what Stan
ultimately utilizes, exhibits much more reasonable tails than the Cauchy
distribution,

```{r}
x <- seq(-10, 10, 0.001)
plot(x, exp(-x) / (1 + exp(-x))**2, type="l", col=c_dark_highlight, lwd=2,
     main="Third Alternative", xlab="x_tilde", ylab="Probability Density",
     yaxt='n')
```

This final implementation requires only a sparse Stan program,

```{r}
writeLines(readLines("cauchy_alt_3.stan"))
```

and the fit proceeds without any indications of problems,

```{r}
fit_3 <- stan(file='cauchy_alt_3.stan', seed=4938483,
              warmup=1000, iter=11000)

util$check_all_diagnostics(fit_3)
```

Once again by modifying the geometry of the target distribution we
accurately recover the quantiles of the Cauchy distribution without
the excessive cost of its nominal implementation,

```{r}
util$plot_estimated_quantiles(fit_3, "Third Alternative")
```

# Comparing Performance of the Various Implementations

To quantify how much better these alternative implementations perform
let's consider the only performance metric that matters for Markov
chain Monte Carlo: the number of effective samples per computational
cost.  Here we use run time as a proxy for computational cost and
consider the effective sample size per time for the indicator function
constructed in each of our models which, unlike the independent
Cauchy components, has a finite mean and variance and hence well-defined
effective sample size.

```{r}
r_nom <- (summary(fit_nom, probs=NA)$summary)[51,5] /
         sum(get_elapsed_time(fit_nom)[,2])
r_1 <- (summary(fit_1, probs=NA)$summary)[151,5] /
        sum(get_elapsed_time(fit_1)[,2])
r_2 <- (summary(fit_2, probs=NA)$summary)[151,5] /
         sum(get_elapsed_time(fit_2)[,2])
r_3 <- (summary(fit_3, probs=NA)$summary)[101,5] /
        sum(get_elapsed_time(fit_3)[,2])

plot(1:4, c(r_nom, r_1, r_2, r_3), type="l", lwd=2, col=c_dark,
     main="", xaxt = "n", xlab="", ylab="ESS / Time (s)",
     xlim=c(0, 5), ylim=c(0, 4000))
axis(1, at=1:4, labels=c("Nom", "Alt 1", "Alt 2", "Alt 3"))
```

Immediately we see that the alternative implementations are
drastically better than the nominal implementation.  Moreover,
the second alternative implementation is almost twice as fast
as the the first.  Perhaps surprisingly, using an inverse Gamma
distribution to avoid a division ends up being a significant
improvement.

We can also compare the geometry of these implementations by
considering only the number of gradient evaluations per iteration
and ignoring the actual cost of each evaluation.


```{r}
probs <- c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9)

sampler_params <- get_sampler_params(fit_nom, inc_warmup=FALSE)
n_grad_nom <- quantile(do.call(rbind, sampler_params)[,'n_leapfrog__'],
                       probs = probs)

sampler_params <- get_sampler_params(fit_1, inc_warmup=FALSE)
n_grad_1 <- quantile(do.call(rbind, sampler_params)[,'n_leapfrog__'],
                     probs = probs)

sampler_params <- get_sampler_params(fit_2, inc_warmup=FALSE)
n_grad_2 <- quantile(do.call(rbind, sampler_params)[,'n_leapfrog__'],
                     probs = probs)

sampler_params <- get_sampler_params(fit_3, inc_warmup=FALSE)
n_grad_3 <- quantile(do.call(rbind, sampler_params)[,'n_leapfrog__'],
                     probs = probs)

idx <- c(1, 1, 2, 2, 3, 3, 4, 4)
x <- c(0.5, 1.5, 1.5, 2.5, 2.5, 3.5, 3.5, 4.5)

cred <- data.frame(n_grad_nom, n_grad_1, n_grad_2, n_grad_3)
pad_cred <- do.call(cbind, lapply(idx, function(n) cred[1:9,n]))

plot(1, type="n", main="", xaxt = "n", xlab="",
     ylab="Gradient Evaluations Per Iteration",
     xlim=c(0, 5), ylim=c(0, 8000))
axis(1, at=1:4, labels=c("Nom", "Alt 1", "Alt 2", "Alt 3"))

polygon(c(x, rev(x)), c(pad_cred[1,], rev(pad_cred[9,])),
        col = c_light, border = NA)
polygon(c(x, rev(x)), c(pad_cred[2,], rev(pad_cred[8,])),
        col = c_light_highlight, border = NA)
polygon(c(x, rev(x)), c(pad_cred[3,], rev(pad_cred[7,])),
        col = c_mid, border = NA)
polygon(c(x, rev(x)), c(pad_cred[4,], rev(pad_cred[6,])),
        col = c_mid_highlight, border = NA)
lines(x, pad_cred[5,], col=c_dark, lwd=2)
```

As expected the alternative implementations require far fewer
gradient evaluations per iteration as they don't exhibit the
heavy tails which require increasingly longer and longer trajectories.

# The Half Cauchy Distribution

These alternative implemenations can also be modified to work for
the half Cauchy distribution over positive values.  In particular,
for the first and second alternative implemenations we just have
to constrain $x_{a}$ to be positive.  In the third we just have
to set
$$
x = \tan \, \left(\frac{\pi}{2} \tilde{x} \right)
$$
to yield a $\text{Half-Cauchy}(0, 1)$ distribution.

To demonstrate let's compare a nominal half Cauchy implementation,

```{r}
writeLines(readLines("half_cauchy_nom.stan"))

fit_half_nom <- stan(file='half_cauchy_nom.stan', seed=4938483,
                     warmup=1000, iter=11000)

util$check_all_diagnostics(fit_half_nom)
```

to the second alternative implementation,

```{r}
writeLines(readLines("half_cauchy_alt.stan"))

fit_half_reparam <- stan(file='half_cauchy_alt.stan', seed=4938483,
                         warmup=1000, iter=11000)

util$check_all_diagnostics(fit_half_reparam)
```

Comparing the 10th identical, Cauchy distribution component of
our model we see that the two implementations yield equivalent
results,

```{r}
x <- extract(fit_half_nom)$x[,10]
p1 <- hist(x[x < 25], breaks=seq(0, 25, 0.25), plot=FALSE)
p1$counts <- p1$counts / sum(p1$counts)

x <- extract(fit_half_reparam)$x[,10]
p2 <- hist(x[x < 25], breaks=seq(0, 25, 0.25), plot=FALSE)
p2$counts <- p2$counts / sum(p2$counts)

c_light_trans <- c("#DCBCBC80")
c_light_highlight_trans <- c("#C7999980")
c_dark_trans <- c("#8F272780")
c_dark_highlight_trans <- c("#7C000080")

plot(p1, col=c_dark_trans, border=c_dark_highlight_trans,
     main="", xlab="x[10]", yaxt='n', ylab="")
plot(p2, col=c_light_trans, border=c_light_highlight_trans, add=T)
```

# Conclusion

Although the nominal implementation of the Cauchy distribution
frustrates computational algorithms with either bias or slow
execution, there exist multiple alternative implementations that
yield an equivalent distribution without the excessive cost.

The practical consequences of these implementations, however,
are ultimately not all that significant.  Having a Cauchy
distribution in the generative model does not necessarily
imply that the heavy tails will persist into the posterior.
In most cases even a little bit of data can tame the heavy
tails, resulting in a pleasant posterior geometry regardless
of which implementation of the Cauchy we use.  Nevertheless
it is up the the user to remain vigilant and monitor for
indications of heavy tails and motivation for alternative
implementations.

# Acknowledgements

I thank Aki Vehtari and Junpeng Lao for helpful comments.

# Original Computing Environment

```{r, comment=NA}
writeLines(readLines(file.path(Sys.getenv("HOME"), ".R/Makevars")))
```

```{r, comment=NA}
devtools::session_info("rstan")
```
