---
title: "Bayes Sparse Regression"
author: "Michael Betancourt"
date: "March 2018"
bibliography: bayes_sparse_regression.bib
output:
  html_document:
    fig_caption: yes
    theme: spacelab #sandstone #spacelab #flatly
    highlight: pygments
    toc: TRUE
    toc_depth: 2
    number_sections: TRUE
    toc_float:
      smooth_scroll: FALSE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment=NA)
```

As complex measurements convolve meaningful phenomena with more and more
extraneous phenomena, _sparsity_ is becoming an increasingly prevalent objective
in statistical analyses.  Much of this zeitgeist has been driven by the success
of frequentist methods like compressed sensing and LASSO regression, and it is
often naively assumed that these properties immediately carry over to the
corresponding Bayesian analyses.  Sparsity in a Bayesian analysis, however, is
induced by fundamentally different properties than those that induce sparsity in
a frequentist analysis.

In this case study I'll review how sparsity arises in frequentist and Bayesian
analyses and discuss the often subtle challenges in implementing sparsity in
practical Bayesian analyses.

# Fading into Irrelevance

The usual setting for sparsity is a regression over a data set containing
many covariates that could possibly correlate with an outcome variate of
interest.  In this circumstance sparsity is the assumption that only a few of
these covariates have any meaningful correlation, although a priori we do not
know exactly which covariates will be relevant and which will be irrelevant.
A sparse regression encodes this assumption to allow the data to inform both
which covariates are relevant and how the relevant covariates then correlate
with the outcome variate.

For example we might be interested in classifying individuals in a population
into two groups, with many individual characteristics possibly influencing the
probabilty of being associated with each group.  Without any sparsity
assumptions the uncertainty in the irrelevant characterisics will propagate
to large uncertainties in inferred associations.  If we can isolate only the
relevant covariates, however, then we can significantly reduce the
uncertainties in how the covariates influence the classification.

For simplicity I will restrict the discussion to only general linear models
where slopes, $\boldsymbol{\beta}$, control the influence of each covariate
through some effective parameter for each observation,
$$
\theta_{n} (\mathbf{x}_{n}, \boldsymbol{\beta}, \alpha)
=
f ( \mathbf{x}_{n} \cdot \boldsymbol{\beta} + \alpha ).
$$
Because a covariate will no longer influence the model when the corresponding
slope vanishes, sparsity is induced by regularizing the irrelevant slopes
towards zero without affecting the relevant slopes and compromising the
resulting inferences

How this sparsity is realized depends intimately on the inferential perspective
that we consider.

## Inducing Sparse Decisions

_Frequentist sparsity_ is a decision making process that explicitly selects
a small subset of covariates to inform the regression while discarding the rest.
When executed properly this selection process can drastically reduce the size
of the data that needs to be considered and hence the computational cost of
making predictions for new observations.  When executed improperly the selection
process can undermine our ability to infer the variate outcome at all.

In the general linear model setting the selection of covariates is equivalent
to identifying which slopes are zero and which are non-zero.  Methods like the
LASSO [@Tibshirani:1996] achieve sparsity by strongly penalizing the maximum
likelihood estimator of the slopes towards small values, and then explicitly
forcing a slope to zero when it falls below a given threshold during the maximum
likelihood estimation.  Any slope that isn't needed to maintain the maximum
likelihood fit to the observed data is forced to zero as the optimization
proceeds and ultimately discarded.

For sufficiently simple data generating processes and large enough data sets
these methods tend to be reasonably well-calibrated.  We will, on average,
discard most of the irrelevant covariates while retaining most of the relevant
covariates, and our abilty ot model the variate outcome, regardless of the
exact data that we observe.

## Inducing Sparse Inferences

_Bayesian sparsity_ avoids decisions altogether and instead manifests as a
distinct behavior in the posterior distribution.  In particular, Bayesian
sparsity arises from a posterior distribution that strongly concentrates around
the neighborhood in parameter space where the irrelevant slopes are zero and
hence inconsequential when making inferences.

As a penalty function can induce sparse decisions in the frequentist setting,
the prior distribution can induce sparse inferences in the Bayesian setting.
But how exactly can a prior distribution coerce the desired concentration
when we don't know which slopes are irrelevant a priori?

It's tempting to appeal to the penalty function that is so critical to the
success in the frequentist setting.  In particular, if we reinterpret a
sparsity-inducing penalty function as a log probability density over parameter
space then does that always define a sparsity-inducing prior distribution?
Unfortunately it does not.  In fact the implied prior distibution can stretch
the corresponding posterior distribution _away_ from the desired neighborhood
where the irrelevan slopes vanish.

The problem is that the sparsity-inducing penalty function has to influence only
a single point in parameter space at any given time, whereas a
sparsity-inducing prior distribution has to consider the entire parameter space
at once.

## Sparsity-Inducing Estimators verses Sparsity-Inducing Distributions

To highlight the difference between inducing sparsity in a point estimator and
inducing sparsity in an entire distribution, let's consider the $L_{1}$ penalty
at the heart of the LASSO,
$$
R_{L_{1}} ( \boldsymbol{\beta} ) =
\sum_{m = 1}^{M} \lambda_{m} \left| \beta_{m} \right|.
$$

When the maximum likeihood estimate of the slope $\beta_{m}$ falls below the
scale $\lambda_{m}$ it is regularized towards zero but, because the penalty is
nearly flat above the scale, estimates above $\lambda_{m}$ experience negligible
regularization.  Given a suitable choice of scales for each covariate this
dichotomous behavior of the penalty facilitates the suppresion of the irrelevant
slopes below the selection threshold while leaving the relevant slopes
undisturbed.

Interpreted as a negative log density, the $L_{1}$ penalty function implies
independent Laplace priors for each of the slopes,
$$
\pi( \boldsymbol{\beta} )
=
\exp (- R_{L_{1}} ( \boldsymbol{\beta} ))
=
\prod_{m = 1}^{M} \exp (- \lambda_{m} \left| \beta_{m} \right| ).
$$
With this prior the mode of the resulting posterior distribution will coincide
with the penalized maximum likelihood estimator; unfortunately the mode is not
a well-posed inference drawn from the posterior distribution.  Proper inferences
correspond instead of posterior expectation values that are informed by the
_entire_ posterior distribution.  The affect of the Laplace prior on the full
posterior distribution is not nearly as useful as its affect on the mode.

Because the maximum likelihood estimator considers only a single point in
parameter space at a time, it is influenced by the either the regularizing
behavior of the penalty below each $\lambda_{m}$ or the laissez faire behavior
above each $\lambda_{m}$, but not both.  The expanse of the posterior
distribution, however, is influenced by both of these behaviors _at the same
time_.  While the shape of the Laplace prior below $\lambda_{m}$ does induce
some concentration of the posterior towards smaller values of $\beta_{m}$, the
heavy tail also drags significant posterior probabilty far above $\lambda_{m}$.

These opposing behaviors induce regrettable features in the posterior for
both the irrelevant slopes, which leak significant probability mass towards
undesired large values, and relevant slopes, which are overshrunk towards
smaller values.  The former prevents us from identifying the irrelevant
parameters while the latter biases our inferences.  Moreover, in both cases the
marginal posteriors become much more diffuse than warranted which inflates our
inferential uncertainties.

If we want to induce sparsity in Bayesian decisions then we need a prior
distribution with a more holistic approach to regularization.  Luckily enough,
we have the horseshoe.

# Constructing Prior Distributions That Induce Sparsity

To implement Bayesian sparsity we need a prior distribution that allows the data
to collapse the entire marginal posterior for each slope towards relevance or
irrelevance, but not both.  In doing so we need to carefully define exactly what
values of the slopes are consequential to the final inference and what values
are not.

## Sparsity-Inducing Shapes

In order to provide the desired flexibility in the posterior for each slope we
need a prior distribution that enforces a global scale while also giving each
of slopes the flexibilty to transcend that scale as needed.  Because we don't
know which slopes will need that flexibilty the desired prior will have to be
exchangeable with respect to the slopes and hence manifest a hierachical
structure.

The _horseshoe_ prior [@CarvalhoEtAl:2009] accomplishes this flexibility by
setting the scale for each component to the product of a global scale, $\tau$,
and a local scale, $\lambda_{m}$, each of which are themselves unknown
parameters,
$$
\begin{align*}
\beta_{m} &\sim \mathcal{N} (0, \tau \cdot \lambda_{m})
\\
\lambda_{m} &\sim \text{Half-}\mathcal{C} (0, 1)
\\
\tau &\sim \text{Half-}\mathcal{C} (0, \tau_{0}).
\end{align*}
$$
The heavy-tailed Cauchy prior distribution for the local scales allows the data
to push each to large values as needed, which then push the corresponding slopes
above the global scale, $\tau$.  By making the global scale itself a
parameter we also allow the data to refine the scale beyond our prior
judgement of $\tau_{0}$.

An immediate problem with the horseshoe prior is that the slopes that transcend
the global scale are otherwise unregularized, leaving their posteriors to
diffuse to extremely large values.  This behavior allows nonidentification or
weak identification of the likelihood to propagate to the posterior and
compromise the validity of the resulting inferences.  The _Finnish horseshoe_
[@PiironenEtAl:2017a] remedies this vulnerability by introducing another level
to the prior hierarchy,
$$
\begin{align*}
\beta_{m} &\sim \mathcal{N} (0, \tau \cdot \tilde{\lambda}_{m})
\\
\tilde{\lambda}_{m} &=
\frac{c \lambda_{m}}
{\sqrt{ c^{2} + \tau^{2} \lambda_{m}^{2}}}
\\
\lambda_{m} &\sim \text{Half-}\mathcal{C} (0, 1)
\\
c^{2} &\sim \text{Inv-}\mathcal{G} \, (\frac{\nu}{2}, \frac{\nu}{2} s^{2})
\\
\tau &\sim \text{Half-}\mathcal{C} (0, \tau_{0}).
\end{align*}
$$
Integrating the new scale, $c$, out of the distribution implies a marginal
$\text{Student-}t \, (\nu, 0, s)$ prior for each of the slopes, at least
sufficiently far above the global scale.  Setting $\nu$ and $s$ appropriately
then ensures that we can contain with the posterior within a few multiples of
$s$ around zero.

## Sparsity-Inducing Thresholds

With the shape of a sparsity-inducing prior established we are left only with
determining the prior hyperparameter $\tau_{0}$ which effetively determines
the scale below which slopes are irrelevant to the modeling of the output
variate.  The subtlety with specifying $\tau_{0}$ is that irrelevance is
determined not by the prior distribution itself but rather our
_measurement process_ -- the contribution of a slope is negligible only when it
is indistinguishable from the inherent variabiltiy of our observations.  As
always the consequences of the prior depend on the context of the likelihood
[@GelmanEtAl:2017].

Let's presume that we will model our measurement process as a linear
regression,
$$
y_{n} \sim
\mathcal{N} (\mathbf{x}_{n} \cdot \boldsymbol{\beta} + \alpha, \sigma).
$$
The measurement variability, $\sigma$, provides a natural scale for irrelevance
in this model.

If we assume that the covariates are independent and distributed around zero
with unit variation then a slope on the order of $\sigma$ will yield an
individual response, $x_{nm} \beta_{m} \approx \sigma$, that would be hard to
discriminate from the expected variation of the measurement itself a priori.
Consequently $\tau_{0} = \sigma$ seems a reasonable assumption.

When we consider the posterior behavior, however, we have to recognize that the
data will typically inform our inferences beyond the scale of the measurement
variability -- with $N$ independent observations we will be sensitive to effects
as small as $\sigma / \sqrt{N}$.  If we ignore the number of observations then
the consequnces of the resulting prior will change with the size of the data!
This suggests instead that we take $\tau_{0} = \sigma / \sqrt{N}$.

Our logic so far has relied on the expected contribution from each slope
individually.  That tails of the horseshoe priors, however, are heavy enough
that when we consider the contribution from many slopes at once we will see
at least a few will exceed the irrelevance threshold and significantly effect
the variate-covariate relationship.  In order to avoid this ensemble behavior we
have to reduce our prior scale even further to
$$
\tau_{0} = \frac{m_{0}}{M - m_{0}} \frac{\sigma}{\sqrt{N}},
$$
where $m_{0}$ is the expected number of relevant slopes.  The sparsity induced
by the corresponding threshold is only weakly sensitive to $m_{0}$, so a
reasonable estimate is typically sufficient.

If we move away from pure linear regression then this argument has to be
modified to account for nonlinearities in the measurement process.
@PiironenEtAl:2017a derives approximate scales appropriate to general linear
models.  In order to faciliate the optimal performance of the horseshoe family
of prior distributions we must take care to ensure that the prior scale is
compatible with the measurement process.

# Experiments

To demonstrate the subtleties in implementing sparsity in a Bayesian analysis
let's consider data simulated from a linear regression model,
```{r}
writeLines(readLines("generate_data.stan"))
```
The covariates are all independently distributed around zero with unit variance,
and there is a population of both large, relevant slopes and small, irrelevant
slopes.  Moreover, the data are collinear, with more covariates than
observations, which implies a non-identified likelihood.

We begin by simulating a measurement,

```{r}
library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

util <- new.env()
source('stan_utility.R', local=util)
source('plot_utility.R', local=util)

fit <- stan(file='generate_data.stan', iter=1,
            chains=1, seed=194838, algorithm="Fixed_param")

X <- extract(fit)$X[1,,]
y <- extract(fit)$y[1,]
N <- dim(X)[2]
M <- dim(X)[1]
beta_true <- extract(fit)$beta[1,]

stan_rdump(c("N", "M", "X", "y", "beta_true"), file="linear_regression.data.R")
```

Only 7 of the simulated slopes are large enough to significantly effect
the regression, with the other slopes left largely irrelevant.

```{r}
input_data <- read_rdump("linear_regression.data.R")

c_dark <- c("#8F2727")
c_dark_highlight <- c("#7C0000")

par(mar = c(4, 4, 0.5, 0.5))
hist(input_data$beta_true, main="", col=c_dark, border=c_dark_highlight,
     xlab="True Slopes", yaxt='n', ylim=c(0, 40), ylab="",
     breaks=11*(-100:100)/100)
```

Our ability to recover these relevant slopes will depend intimately on the prior
we incorporate into our Bayesian model.

## Uniform Prior

Let's begin with a uniform prior over the slopes,
```{r}
writeLines(readLines("linear_regression_unif.stan"))
```

```{r, cache=TRUE}
unif_fit <- stan(file='linear_regression_unif.stan',
                 data=input_data, seed=4938483)
```

The uniform prior allows that non-identifiabilty of the likelihood to propagate
to the posterior.  The fit of the resulting posterior unsurprisingly fails in
spectacular fashion, with vanishing effective sample sizes, large $\hat{R}$, and
failing HMC diagnostics.

```{r}
capture.output(util$check_n_eff(unif_fit))[1:5]
capture.output(util$check_rhat(unif_fit))[1:5]
util$check_div(unif_fit)
util$check_treedepth(unif_fit)
util$check_energy(unif_fit)
```

The fit does, however, exhibit some of the massive uncertainty inherent to
the non-identified posterior.  This is evident from looking at the extreme
extent of the marginal posteriors for each slope, here visualized as nested
intervals ranging from the median in dark red to an 80% posterior interval in
the lightest red.  The relevant slopes are indicated by dashed grey lines and
the truth slopes are shown in black.  Note the scale of the y-axis here
surpassing thousands when the largest slopes shouldn't be much magnitudes much
greater than ten.

```{r}
util$plot_post_quantiles(unif_fit, input_data, "Uniform Prior")
```

Subtracting the true slopes from the inferred slopes, with black now indicating
the desired zero baseline, doesn't show any inaccuracy but only because the
posterior is so non-identified.

```{r}
util$plot_residual_quantiles(unif_fit, input_data, "Uniform Prior")
```

The intercept and measurement variability are better identified, but only
because of their weakly informative priors.

```{r}
util$plot_aux_posteriors(unif_fit, "Uniform Prior")
```

## Narrow Weakly Informative Prior

We definitely need a prior to compensate for the non-identified likelihood,
but just how much prior informationt do we need?  Let's try a weakly-informative
prior for all of the slopes that strongly concentrates below the scale of the
measurement variability.

```{r}
writeLines(readLines("linear_regression_narrow.stan"))
```

```{r, cache=TRUE}
narrow_fit <- stan(file='linear_regression_narrow.stan',
                   data=input_data, seed=4938483)
```

The fit is now much better behaved,

```{r}
util$check_all_diagnostics(narrow_fit)
```

but the extreme regularization biases the marginal posteriors for the relevant
slopes, again shown with grey dashed lines, far below their true values, again
in black.

```{r}
util$plot_post_quantiles(narrow_fit, input_data, "Narrow Prior")
```

This is even more evident if we subtract the true values.

```{r}
util$plot_residual_quantiles(narrow_fit, input_data, "Narrow Prior")
```

Additionally, the inferred measurement variability is biased quite high in
order to compensate for the underestimated relevant slopes.

```{r}
util$plot_aux_posteriors(narrow_fit, "Narrow Prior")
```

## Wide Weakly Informative Prior

Would a weakly informative prior work if we expanded the scale so to
encompass the breadth of the true slopes?

```{r}
writeLines(readLines("linear_regression_wide.stan"))
```

```{r, cache=TRUE}
wide_fit <- stan(file='linear_regression_wide.stan',
                 data=input_data, seed=4938483)
```

The fit diagnostics hint at trouble, which is not unexpected given that the
wider prior allows more of the likelihood non-identifiabilty to spread to the
posterior.

```{r}
util$check_all_diagnostics(wide_fit)
```

More importantly, the wide prior offers little regularization to the many slopes
that are expected to be negligible.  The resulting marginal posteriors exhibit
little contraction away from the prior.

```{r}
util$plot_post_quantiles(wide_fit, input_data, "Wide Prior")

util$plot_aux_posteriors(wide_fit, "Wide Prior")
```

The residuals don't demonstrate any significant inaccuracy, but only because
of the large uncertainties.

```{r}
util$plot_residual_quantiles(wide_fit, input_data, "Wide Prior")
```

## Laplace Prior

If we want to isolate the relevant slopes then we have to encode the
assumption of sparsity into our prior distribution.  Before considering
the family of horseshoe priors, however, let's take a look at the Laplace
prior motivated by a naive translation of the frequentist LASSO.

```{r}
writeLines(readLines("linear_regression_laplace.stan"))
```

```{r, cache=TRUE}
laplace_fit <- stan(file='linear_regression_laplace.stan',
                    data=input_data, seed=4938483,
                    control=list(adapt_delta=0.99, max_treedepth=12))
```

Even with the large treedepths and small adaptation target, `adapt_delta`, the
fit with the Laplace prior exhibits slow mixing.

```{r}
util$check_n_eff(laplace_fit)
util$check_rhat(laplace_fit)
util$check_div(laplace_fit)
util$check_treedepth(laplace_fit, 12)
util$check_energy(laplace_fit)
```

The Laplace prior finally yields some of the behavior that we need to encode
sparsity, resulting in much better behavior compared to the failures up to
this point.  Still, the relevant slopes exhibit signs of overregularization
while the irrelvant slopes aren't as strongly regularized as we'd like.

```{r}
util$plot_post_quantiles(laplace_fit, input_data, "Laplace Prior")

util$plot_aux_posteriors(laplace_fit, "Laplace Prior")
```

This is particularly evident once we subtract the true slopes from the recovered
posteriors.

```{r}
util$plot_residual_quantiles(laplace_fit, input_data, "Laplace Prior")
```

## The Horseshoe

In order to encode proper inferential sparsity into our Bayesian analysis we
need to consider the horseshoe or variants thereof.  Let's first consider the
classic horseshoe with the often recommended scale $\tau_{0} = \sigma$.

```{r}
writeLines(readLines("linear_regression_horseshoe.stan"))
```

```{r, cache=TRUE}
horseshoe_fit <- stan(file='linear_regression_horseshoe.stan',
                      data=input_data, seed=4938483,
                      control=list(adapt_delta=0.99, max_treedepth=15))
```

Unfortunately, the fit isn't great.  The horseshoe does have a reputation for
being challenging to fit in general, but its lack of regularization of the
relevant slopes is particularly troublesome in the context of the
non-identified likelihood.

```{r}
util$check_n_eff(horseshoe_fit)
util$check_rhat(horseshoe_fit)
util$check_div(horseshoe_fit)
util$check_treedepth(horseshoe_fit, 15)
util$check_energy(horseshoe_fit)
```

Although the fit isn't trustworthy, we do see some sparsity manifesting in the
marginal posteriors for the irrelevant slopes.

```{r}
util$plot_post_quantiles(horseshoe_fit, input_data, "Horseshoe Prior")
```

Unfortunately the stress of the non-identified likelihood prevents the horseshoe
prior from enabling accurate fit of the relevant slopes.

```{r}
util$plot_residual_quantiles(horseshoe_fit, input_data, "Horseshoe Prior")
```

The marginal posterior for the measurement variability, $\sigma$, also
exhibits the poor mixing of the fit.

```{r}
util$plot_aux_posteriors(horseshoe_fit, "Horeshoe Prior")
```

## The Finnish Horseshoe

Finally we can consider incorporating a Finnish horseshoe into our model.
Here we assume 10 large slopes and tune the extra level of regularization to
place most of the prior mass below a magnitude of 10.

```{r}
writeLines(readLines("linear_regression_finnish_horseshoe.stan"))
```

As with any Finn, the Finnish horseshoe can be a bit stubborn and we have to
push the limits of Hamiltonian Monte Carlo a bit to resolve all of the
structure it induces in the posterior distribution.  With some tuning, however,
we recover a good fit.

```{r, cache=TRUE}
finnish_fit <- stan(file='linear_regression_finnish_horseshoe.stan',
                    data=input_data, seed=4938483,
                    control=list(adapt_delta=0.99, max_treedepth=15))
```

```{r}
util$check_n_eff(finnish_fit)
util$check_rhat(finnish_fit)
util$check_div(finnish_fit)
util$check_treedepth(finnish_fit, 15)
util$check_energy(finnish_fit)
```

The Finnish horseshoe induces narrow posteriors for both the relevant and
irrelevant slopes without compromising accuracy.

```{r}
util$plot_post_quantiles(finnish_fit, input_data, "Finnish Horseshoe Prior")
```

In particular, the marginal posterior residuals show no sign of a poor fit.

```{r}
util$plot_residual_quantiles(finnish_fit, input_data, "Finnish Horseshoe Prior")
```

```{r}
util$plot_aux_posteriors(finnish_fit, "Finnish Horeshoe Prior")
```

## Comparison

To emphasize the relative performance of the candidate sparsity-inducing
priors let's compare the marginal posterior residuals for only the slopes,
$\left\{ \beta_{51}, \ldots, \beta_{100} \right\}$, using the same plotting
ranges.

As expected the Laplace prior achieves some inferential sparsity in the
posterior but the relevant slopes show signs of overregularization and the
irrelevant slopes show signs of over overregularization.

```{r}
util$plot_summary_residual_quantiles(laplace_fit, input_data,
                                     "Laplace Horseshoe Prior")
```

Because of the non-identified likelihood the horseshoe prior yields a complex
posterior that Stan isn't able to accurately fit, but what we do recover
exhibits extreme overregularization of the relevant slopes.

```{r}
util$plot_summary_residual_quantiles(horseshoe_fit, input_data,
                                     "Horseshoe Prior")
```

By regularizing the extent of the relevant slopes in addition to the irrelevant
slopes, the Finnish horseshoe ensures an identified posterior that exhibits
exactly the dichotomous regularization we need to properly induce sparsity.

```{r}
util$plot_summary_residual_quantiles(finnish_fit, input_data,
                                     "Finnish Horseshoe Prior")
```

# Conclusion

Sparsity is a potentially powerful but ultimately quite subtle assumption.
Only by utilizing prior distributions that induce sparsity in a principled
manner can we truly exploit the assumption of sparsity in those circumstances
where it is appropriate.  In particular, any proper application of sparsity
requires a detailed understanding of how the sparse parameters effect the
measurement process.

Finally, sparse inferences and sparse decisions are not mutually exclusive.
Indeed inferential sparsity is critical for enabling robust sparse decisions
in a Bayesian framework.  For example, @PiironenEtAl:2017b use
inferential sparsity to faciliate variable selection that minimizes the loss
of predictive performance.

# Acknowledgements

I thank Dan Simpson, Aki Vehtari, and Juho Piironen for many helpful
conversations about sparsity and comments on this case study.

# Original Computing Environment

```{r, comment=NA}
writeLines(readLines(file.path(Sys.getenv("HOME"), ".R/Makevars")))
```

```{r, comment=NA}
devtools::session_info("rstan")
```

# References
